Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-11 Ben_Manuscripts/hdphmm/Draft.tex
comments on draft.

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  %BJC3: I emailed Emily Fox and she just asked me to use this citation. I'm happy to add a 
  %link to where she stores the code. Would that be best in the form of a reference?
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  We detected equilibration of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

@@

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by cylindrical 
  coordinates while estimating the state sequence.
  %BJC3: maybe could say something about multivariate guassian noise with VAR in cylindrical doesnt work with current code.

@@

@@

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. We 
  converted the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ as $\lambda_x$ + $\lambda_y$
  and $\lambda_x^2$ + $\lambda_y^2$ respectively. %BJC3: possibly could use some justification of these definitions. 
  We obtained means in $r$, $\mu_r$, using the norm of the $x$ and $y$ means, 
  $\sqrt{\mu_x^2 + \mu_y^2}$. We squared $\mu_r$ for clustering in order to cast it in the
  same units as $A$ and $\Sigma$.

@@

  %BJC3: modified
  We remapped and fixed the state sequence based on the cluster assignments and
  parameterized the clusters using the inference component of the IHMM algorithm. 
  First, we modified the ($x$, $y$, $z$) solute trajectories so that they had a 
  mean of zero, leaving only the fluctuations. First, we zeroed the $y$ component
  of the trajectories by rotating the coordinates about the pore axis such that 
  the mean vector of each separate segment was aligned with the $x$ axis. Next, 
  we subtracted the means from each separate segment in the $x$ and $z$ dimensions.
  We used the IHMM algorithm on this modified trajectory to infer the clustered 
  state parameters by fixing the clustered state sequence.

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
  each value of the rotated $\mathbf{c}$ vectors assigned to the same cluster. Note
  that we only care about the $x$ and $y$ component of $\mathbf{c}$ because solute 
  trajectories are not bound in the $z$ direction.

@@

  %MRS2: I'd make this a separate subsection ``Deviations of the IMMM from molecular motion'' or something like that.
  % BJC3: would these subsubsection headers look better as just italicized and on 
  % the same line as where the next paragraph starts?
  \subsubsection*{Deviations of the IHMM from Molecular Motion}

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{hbonds_assoc_summary.pdf}
  \caption{}\label{fig:hbonds_assoc_summary}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{rdf_summary.pdf}
  \caption{}\label{fig:rdf_summary}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{dwell_time_summary.pdf}
  \caption{}\label{fig:dwell_time_summary}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{cov_summary.pdf}
  \caption{}\label{fig:cov_summary}
  \end{subfigure}
  %MRS7: is c) methanol?  I thought it wasn't associating with Na+ much.
  %BJC7: looks about 5 %, certainly less than the other. I could add
  % some discussion of this to previous section.
  %MRS: methanol yellow is hard to see in (b). Maybe add outlines to histogram?  I guess they are, increase alpha or something for yello?
  %MRS: check consistency of capitalization in legend in (c) 
  \caption{(a) Solutes hydrogen bond and associate with sodium ions
  to different degrees. Sometimes, they participate in both interactions at
  once. The frequency at which solutes interact via either mechanism does
  not appear to be linearly related to the solute's mean squared displacement
  after a 1000 ns time lag. (b) We histogrammed the means of the clusters weighted
  by the number of fluctuation within that cluster. Most solutes spend the majority of
  their time close to the pore center. Only methanol has a significant peak
  beyond 1.5 nm from the pore center. (c) The trend in the maximum dwell time
  among prevalent states for each solute is inversely related to the
  solute's MSD. In some cases, non-prevalent states show very long dwell times,
  which may have a significant impact on solute MSD. (d) All solutes tend to
  take larger hops in the $z$ direction when they are in the pore versus in 
  the tails. All solutes except urea make slightly larger radial hops while 
  in the tails.
  %BJC6: I made this plot out of curiosity, not sure if I'll use it yet.
  }\label{fig:summaries}
  \end{figure}

@@

% LocalWords:  BJC micropollutants Desalination permeability nanofiltration LLC
% LocalWords:  Amphiphilic nanostructures Lyotropic amphiphilic lyotropic MSDDM
% LocalWords:  solutes selectivities solute nanoscopic timescales MSDs MSD HMMs
% LocalWords:  nonparameteric Parrinello Rahman barostat rescale transitioning
% LocalWords:  al's HMM Dirichlet HDP DP Wishart TODO SI al MATLAB teh gael pdf
% LocalWords:  nonparametric bayesian solute's reparameterized intracluster ca
% LocalWords:  delocalize Luzar Chandler histogrammed interpolator methanol's
% LocalWords:  nanostructure carboxylate VMD nonidealities hbonds outlier GAANN
% LocalWords:  autogregressive PRF ACS XSEDE ACI

>>>>

  Using the IHMM framework, we estimated the most likely sequence of hidden states in
  each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  We created a Python implementation of this process which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_bayesian_2010} Parameter estimation is iterative. 
  %MRS2: give location for the code here where you introduce it.
  %BJC3: I emailed Emily Fox and she just asked me to use this citation. I'm happy to add a 
  %link to where she stores the code. Would that be best in the form of a reference?
  %MRS3: I was thinking of citing YOUR code here.
  Therefore, we looked for convergence of the parameters in Equation~\ref{eqn:var}.
  We detected equilibration of the parameters using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of parameter estimates. Plots illustrating parameter convergence are 
  available in Section~\ref{S-section:convergence} of the Supporting Information. We 
  refer the interested reader to much more extensive descriptions of the inference and 
  sampling procedures used to estimate the VAR(1) parameters and the state sequence. ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}

@@

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center. We tracked the solute's
  motion along the pore axis with the center-of-mass $z$ coordinate. Using the nearest
  pore center as the origin, we represented the radial distance of each solute's 
  center-of-mass from the pore center in 2 dimensions, $x$ and $y$. By working in 
  Cartesian coordinates, we avoid mathematical complexity introduced by change of coordinates into cylindrical 
  coordinates while estimating the state sequence.
  %BJC3: maybe could say something about multivariate Gaussian noise with VAR in cylindrical doesn't work with current code.
  %MRS3: tweaked above, probably OK.

@@

  %MRS3: I think these belong after clustering.

@@

  To take advantage of the system's cylindrical symmetry, we converted the parameters that
  are directly related to solute position into cylindrical coordinates before clustering. 
  %MRS3: tweaked: 
  %We converted the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ as $\lambda_x$ + $\lambda_y$
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with
  $\lambda_x$ + $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$
  respectively, which are invariant with rotation around the $z$ axis,
  in both cases reducing the dimensionality from 3 to 2.  We clustered
  on squared means in $r$, $\mu_r^2 = \mu_x^2 + \mu_y^2$, rather than
  $\mu_x$ and $\mu_y$ individually, with the square $\mu_r$ converting
  it into the same units as $A$ and $\Sigma$.

@@

  %BJC3: modified

  We remapped and fixed the state sequence based on
  the cluster assignments and parameterized the clusters using the
  inference component of the IHMM algorithm.  First, we modified the
  ($x$, $y$, $z$) solute trajectories so that they had a mean of zero,
  leaving only the fluctuations. Next, we took advantage of the
  rotational invariance of the system by rotating the
  coordinates about the pore axis such that the mean vector of each
  separate segment was aligned with the $x$ axis. Next, we subtracted
  the means from each separate segment in the $x$ and $z$ dimensions,
  with the mean in $y$ being zeroed by the rotation.  We used the IHMM
  algorithm on this modified trajectory to infer the clustered state
  parameters by fixing the clustered state sequence.

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
  each value of the rotated $\mathbf{c}$ vectors assigned to the same cluster. Note
  %MRS: tweaked
  that we only care about the $r=\sqrt{x^2+y^2}$ portion of $\mathbf{c}$ because solute 
  trajectories are not bound in the $z$ direction.

@@

  %MRS2: I'd make this a separate subsection ``Deviations of the IMMM from molecular motion'' or something like that.
  % BJC3: would these subsubsection headers look better as just italicized and on 
  % the same line as where the next paragraph starts?
  %MRS3: yes, that's what I'm thinking of.
  \subsubsection*{Deviations of the IHMM from Molecular Motion}

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{hbonds_assoc_summary.pdf}
  \caption{}\label{fig:hbonds_assoc_summary}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{rdf_summary.pdf}
  \caption{}\label{fig:rdf_summary}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{dwell_time_summary.pdf}
  \caption{}\label{fig:dwell_time_summary}
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
  \includegraphics[width=\textwidth]{cov_summary.pdf}
  \caption{}\label{fig:cov_summary}
  \end{subfigure}
  %MRS7: is c) methanol?  I thought it wasn't associating with Na+ much.
  %BJC7: looks about 5 %, certainly less than the other. I could add
  % some discussion of this to previous section.
  %MRS: methanol yellow is hard to see in (b). Maybe add outlines to histogram?  I guess they are, increase alpha or something for yellow?
  %MRS: check consistency of capitalization in legend in (c) 
  \caption{(a) Solutes hydrogen bond and associate with sodium ions
  to different degrees. Sometimes, they participate in both interactions at
  once. The frequency at which solutes interact via either mechanism does
  not appear to be linearly related to the solute's mean squared displacement
  after a 1000 ns time lag. (b) We histogrammed the means of the clusters weighted
  by the number of fluctuation within that cluster. Most solutes spend the majority of
  their time close to the pore center. Only methanol has a significant peak
  beyond 1.5 nm from the pore center. (c) The trend in the maximum dwell time
  among prevalent states for each solute is inversely related to the
  solute's MSD. In some cases, non-prevalent states show very long dwell times,
  which may have a significant impact on solute MSD. (d) All solutes tend to
  take larger hops in the $z$ direction when they are in the pore versus in 
  the tails. All solutes except urea make slightly larger radial hops while 
  in the tails.
  %BJC6: I made this plot out of curiosity, not sure if I'll use it yet.
  }\label{fig:summaries}
  \end{figure}

@@

% LocalWords:  BJC micropollutants Desalination permeability nanofiltration LLC
% LocalWords:  Amphiphilic nanostructures Lyotropic amphiphilic lyotropic MSDDM
% LocalWords:  solutes selectivities solute nanoscopic timescales MSDs MSD HMMs
% LocalWords:  nonparameteric Parrinello Rahman barostat rescale transitioning
% LocalWords:  al's HMM Dirichlet HDP DP Wishart TODO SI al MATLAB teh gael pdf
% LocalWords:  nonparametric bayesian solute's reparameterized intracluster ca
% LocalWords:  delocalize Luzar Chandler histogrammed interpolator methanol's
% LocalWords:  nanostructure carboxylate VMD nonidealities hbonds outlier GAANN
% LocalWords:  autogregressive PRF ACS XSEDE ACI IHMM nanoporous flowback evy
% LocalWords:  timescale subdiffusive coscia subdiffusion calderon GROMACS der
% LocalWords:  berendsen gromacs spoel hess emailed plateauing hmm ne gelman AR
% LocalWords:  walt numpy GCL URE ACH IMMM subsubsection modT convective vmd Na
% LocalWords:  scannable subsubcaptions errorbars uncharacteristicness facelift
% LocalWords:  PSC RMACC hdphmm TOC
